The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-12-04 Paper.qmd

---
title: "The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study"
author:
  - name: Michael R. Gaffney
    orcid: 0000-0002-7552-6942
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Jessica K. Hlay
    orcid: 0000-0002-6139-5144
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Izabel Rodriguez
  - name: Kristen L. Syme
    orcid: 0000-0002-5601-8897
    corresponding: false
    affiliation:
      - name: Leiden University
        department: Institute of Security and Global Affairs
  - name: Aaron D. Blackwell
    orcid: 000-0002-5871-9865
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Carolyn R. Hodges-Simeon
    orcid: 0000-0002-0691-9080
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Edward H. Hagen
    orcid: 0000-0002-5758-9093
    email: edhagen@wsu.edu
    corresponding: true
    affiliation:
      - name: Washington State University
        department: Anthropology
execute:
  echo: false
  warning: false
  keep-md: true
format:
  docx:
    number-sections: true
    prefer-html: true
  html:
    toc: true
    toc-expand: 2
    number-sections: true
lightbox: false
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

```{r setup, include=FALSE}
source("dictionaries.R", local = knitr::knit_global())
source("analysisforpaper.R", local = knitr::knit_global())

library(modelsummary)
library(english)
# source("regularize.R")
```

@@

To gain support, children use signals to communicate their needs and wants to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has rarely been investigated in a single study of children of all ages. To fill this gap, we collected mother and other primary caregiver reports of child signaling from `{r} nrow(caregivers)` families with `{r} nrow(modeldf)` children on Utila, a small island off the coast of Honduras. In exploratory analyses, we found that child signaling was common in both sexes and across all ages, although it decreased with age and neighborhood quality and increased with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were sad more frequently were perceived as needier within the household and were more likely to receive investment. Consistent with previous research, increased alloparenting effort was associated with more signaling by girls but less signaling by boys. Our results suggest that evolutionary theories of signaling can help explain patterns of child sadness, crying, and temper tantrums.

@@

This study of children aged `{r} paste(range(modeldf$ChildAge), collapse="-")` and their parents and other primary caretakers, was conducted in Utila, Honduras, one of the country’s small Bay Islands in the Western Caribbean. Participants in the study were sampled from two populations who face different levels of economic and social adversity: 1) a long-standing English-speaking community primarily comprising descendants of British immigrants who arrived on the island in the 1800’s, and 2) a more recent community of Spanish-speakers, mostly from mainland Honduras. Compared to the former, the latter are more likely to live in crowded and suboptimal conditions, face active discrimination and increased stress [@garcia2017], and experience economic uncertainty due the predominance of low paying, seasonal jobs primarily related to the island’s diving tourism industry [@garcia2018; @stonich2000]. Such inequality appears to be increasing in response to recent growth in the tourism industry, with the majority of wealth from tourism flowing to a small portion of native English-speaking islanders and other Westerners [@davis2014].

@@

Caretakers were also asked three questions using a five-point Likert scale about their child's susceptibility to illness. We reversed the scores on two measures so that a score of 5 corresponded with the lowest susceptibility to illness before we averaged the scores of these to create a mean *illness avoidance* variable. In the `{r} sum((!is.na(modeldf$IllnessSusceptibility2)) & is.na(modeldf$IllnessSusceptibility3))` instance in which we only had data for two of the measures, we still calculated the mean but used a numerator of 2. See @stbl-healthvars for the wording of these variables.

@@

```{r}
# primary_caregivers <- str_to_sentence(english::english(sum(caregivers$HasSignalingData)))

tbl <- table(caregivers$HasSignalingData, caregivers$CaregiverSex)
primary_caregivers <- tbl[2,1] + tbl[2,2] + tbl[2,3]
```

`{r} primary_caregivers` primary caregivers (females = `{r} tbl['TRUE', 'Female']`, males = `{r} tbl['TRUE', 'Male']`, non-binary = `{r} tbl['TRUE', 'Non-binary / third gender']`) took the survey and provided signaling data for at least one child. `{r} nrow(modeldf)` children had frequency data for at least one signal, with `{r} nrow(modeldf_FULLSIG)` having data for all three. Among those with at least one record of signaling frequency, `{r} sex_table[1]` were female and `{r} sex_table[2]` were male. For summary statistics of study variables see @tbl-sumtab_combined and @tbl-sumtab_factors.

@@

kable(custom.summarize(modeldf, longform_dict[1:23])) #%>% kable_styling()
```

@@

Primary caretakers rarely described a child as “much less” `{r} fmt_sum(modeldf$RelativeNeed == "Much less")` \[fix in word doc?\] or “much more” `{r} fmt_sum(modeldf$RelativeNeed == "Much more")` needy compared to their other children. Therefore, we collapsed “less” and “much less” into a “less” factor and “more” and “much more” into a “more” factor. The end result was `{r} rel_need_table[1]` less needy children, `{r} rel_need_table[2]` children who were as needy as their siblings, and `{r} rel_need_table[3]` more needy children.

@@

>>>>

---
title: "The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study"
author:
  - name: Michael R. Gaffney
    orcid: 0000-0002-7552-6942
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Jessica K. Hlay
    orcid: 0000-0002-6139-5144
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Izabel Rodriguez
  - name: Kristen L. Syme
    orcid: 0000-0002-5601-8897
    corresponding: false
    affiliation:
      - name: Leiden University
        department: Institute of Security and Global Affairs
  - name: Aaron D. Blackwell
    orcid: 000-0002-5871-9865
    corresponding: false
    affiliation:
      - name: Washington State University
        department: Anthropology
  - name: Carolyn R. Hodges-Simeon
    orcid: 0000-0002-0691-9080
    corresponding: false
    affiliation:
      - name: Boston University
        department: Anthropology
  - name: Edward H. Hagen
    orcid: 0000-0002-5758-9093
    email: edhagen@wsu.edu
    corresponding: true
    affiliation:
      - name: Washington State University
        department: Anthropology
execute:
  echo: false
  warning: false
  keep-md: true
format:
  docx:
    number-sections: true
    prefer-html: true
  html:
    toc: true
    toc-expand: 2
    number-sections: true
    embed-resources: true
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
crossref:
  custom:
    - kind: float
      reference-prefix: Figure S
      key: sfig
      space-before-numbering: false
    - kind: float
      reference-prefix: Table S
      key: stbl
      space-before-numbering: false
---

```{r setup, include=FALSE}
if(!dir.exists("Figures")) dir.create("Figures")

source("recode.R")
source("dictionaries.R")

# To use public data, comment out dataprep.R
# and uncomment dataprepPublic.R
source("dataprep.R")
# source("dataprepPublic.R")

source("analysis.R")
source("analysis2.R") # Generates many plots. Slow!
```

@@

To gain support, children use signals to communicate their needs and wants to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations. However, the full range of potential child signals of need, which extend beyond cries, has rarely been investigated in a single study of children of all ages. To fill this gap, we collected mother and other primary caregiver reports of child signaling from `{r} n_distinct(householdIDs)` families with `{r} nrow(utila_df)` children on Utila, a small island off the coast of Honduras. In exploratory analyses, we found that child signaling was common in both sexes and across all ages, although it decreased with age and neighborhood quality and increased with the frequency of conflict between children and caretakers. Consistent with signaling theory, children who were sad more frequently were perceived as needier within the household and were more likely to receive investment. Consistent with previous research, increased alloparenting effort was associated with more signaling by girls but less signaling by boys. Our results suggest that evolutionary theories of signaling can help explain patterns of child sadness, crying, and temper tantrums.

@@

This study of children aged `{r} paste(range(utila_df$ChildAge), collapse="-")` and their parents and other primary caretakers, was conducted in Utila, Honduras, one of the country’s small Bay Islands in the Western Caribbean. Participants in the study were sampled from two populations who face different levels of economic and social adversity: 1) a long-standing English-speaking community primarily comprising descendants of British immigrants who arrived on the island in the 1800’s, and 2) a more recent community of Spanish-speakers, mostly from mainland Honduras. Compared to the former, the latter are more likely to live in crowded and suboptimal conditions, face active discrimination and increased stress [@garcia2017], and experience economic uncertainty due the predominance of low paying, seasonal jobs primarily related to the island’s diving tourism industry [@garcia2018; @stonich2000]. Such inequality appears to be increasing in response to recent growth in the tourism industry, with the majority of wealth from tourism flowing to a small portion of native English-speaking islanders and other Westerners [@davis2014].

@@

Caretakers were also asked three questions using a five point Likert scale about their child's susceptibility to illness. We reversed the scores on two measures so that a score of 5 corresponded with the lowest susceptibility to illness before we averaged the scores of these to create a mean *illness avoidance* variable. In the `{r} sum((!is.na(utila_df$IllnessSusceptibility2)) & is.na(utila_df$IllnessSusceptibility3))` instance in which we only had data for 2 of the measures, we still calculated the mean but used a numerator of 2. See @stbl-healthvars for the wording of these variables.

@@

```{r}
sex_table <- table(utila_df$Sex)
```

`{r} xfun::numbers_to_words(n_distinct(householdIDs), cap=T)` primary caretakers (females = `{r} table(caregiverSex)[1]`, males = `{r} table(caregiverSex)[2]`, missing sex = `{r} sum(is.na(caregiverSex))`) took the survey and provided signaling data for at least one child. `{r} xfun::numbers_to_words(nrow(utila_df), cap=T)` children had frequency data for at least one signal, with `{r} nrow(utila_FULLSIG)` having data for all three. Among those with at least one record of signaling frequency, `{r} sex_table[1]` were female and `{r} sex_table[2]` were male. For summary statistics of study variables see @tbl-sumtab_combined and @tbl-sumtab_factors.

@@

kable(custom.summarize(utila_df, longform_dict[1:23])) %>% kable_styling()
```

@@

```{r}
rel_need_table <- table(utila_df$RelativeNeed3)
```

Primary caretakers rarely described a child as “much less” `{r} fmt_sum(utila_df$RelativeNeed == "Much less")` \[fix in word doc?\] or “much more” `{r} fmt_sum(utila_df$RelativeNeed == "Much more")` needy compared to their other children. Therefore, we collapsed “less” and “much less” into a “less” factor and “more” and “much more” into a “more” factor. The end result was `{r} rel_need_table[1]` less needy children, `{r} rel_need_table[2]` children who were as needy as their siblings, and `{r} rel_need_table[3]` more needy children.

@@

```{r}
response_overlap <- table(utila_df$PositiveResponse, utila_df$NegativeResponse)
```
